Exploring Uncharted Territories - Predicting Activty Cliffs in Structure-Activity Landscapes

### 2012-07-26 rebuttal.tex

\textbf{(3) The derivation of RF models is essentially not described (except of making reference to
  default parameters settings in R). It remains largely unclear how these models were developed. }

The text has been updated to be more specific on the development of the random forest model.

\textbf{(4) The selection of training and test sets should be clearly specified in the Methods section. For
example, how were the RF models trained and evaluated for which predictions of pair-wise SALI values
are reported in Figure 2? It is hoped that the reported SALI values were not predicted for the
training data.}

We have updated the text to note the training/test split and have include results for both sets in
the text. However, due to the similar peformance of models built using the whole dataset (and the
fact that the RF approach trains and tests on out-of-bag data implicitly), we used models built
using the entire dataset (save for some hold out molecules) in the latter half of Section 4.

\textbf{(5) No support is provided for the assumption that the RF models “are relatively robust even for
datasets of small size”. In fact, the author later on states several times that limited model
quality might be due to the use of small compound sets. Appropriate validation will be essential
here.}

We have removed this statement as it is not necessarily a function of the random forest approach,
and more a function of the nature of the dataset. Small datasets, if appropriately constructed can
lead to low RMSE. In our case, we simulated by sampling from the Cavalli pairwise dataset, using
different sample sizes. As shown in the attached figure, model performance is clearly a function of
sample size. Importantly, for certain small samples, the model performance can be quite good. But of
course, this is due to luck.

@@

\textbf{(6) There is only limited statistical assessment/validation of the models. Statements made in the
text that R2 values are “reasonable” and the models perform “relatively well” are difficult to
reconcile. Rather, Table 2 indicates that the R2 values were rather poor in several instances.}

The text has been updated to read more quantitatively and avoid ambiguous statements

\textbf{(7) It appears that there are systematic prediction errors for pairs. At low value ranges (that are
not relevant for activity cliff assessment), SALI values are consistently over-predicted. By
contrast, at high value ranges (that are relevant), SALI values are consistently under-predicted, in
part significantly. This is not investigated, although the apparent errors at high value ranges are
the perhaps most critical features of the reported RF models.}

\textbf{(8) In the section “Extending a landscape”, predictions are finally reported for only three (!)
molecules taken out of each data set. Carefully put, the predictions are heterogeneous. On the basis
of the results reported for this little bit of evaluation, one would be hard pressed to make a case
for the ability of the approach to predict activity cliffs with any certainty.}

### 2012-07-26 paper.tex
Added more edits to tset/pset split

Given the independent and dependent variables for the pairwise dataset, we can now proceed to model
development. For this study we focused on the use of random forest models\cite{Breiman:1984aa}. This
was motivated by the fact that such models tend not to overfit the training data (given a
sufficiently large number of trees) and the fact that the algorithm implicitly performs feature
selection. As a result, this allows us to forgo an explicit feature selection step and work directly
with the descriptor pool (after removal of correlated and constant descriptors). Furthermore, there
is no reason, \emph{a priori} to assume that the underlying SAR's are linear. A random forest model,
being an algorithmic approach\cite{Breiman:2001nx} (as opposed to a distributional one such as
linear regression) makes no such assumptions. We employed the implementation of random forest from
the \texttt{randomForest} package, in R 2.11.0\cite{r}. The models were built using 500 trees, and
randomly sampled the descriptor pool for a given dataset, using $\sqrt(N)$ descriptors at a time,
where $N$ is the number of descriptors in the pool. By default, the method builds individual trees
using 63\% of the dataset and tested on the remainder (the so called out-of-bag data). The final
predicted values for the entire dataset are obtained by averaging the predictions for each
observation from from all 500 trees. We also investigated the traditional approach of a
training/test set split (80\% of the dataset was assigned to the training set) and the results were
similar. As a result, the remainder of this work employed models built using the entire dataset.

@@

\section{Results}
\label{sec:applications}
We first consider the application of the SALI prediction methodology to the Cavalli dataset. We
developed three random forest models, using the three aggregation descriptor aggregation functions
described in Section \ref{sec:methodology}. Fig.~\ref{fig:rfmodel-cavalli} presents the predicted
versus observed SALI values from the three models and Table \ref{tab:rfmodel-results} summarizes the
performance metrics for the three models. Overall the different aggregation functions do not differ
dramatically in terms of the final model performance as root mean square error (RMSE) ranges from
0.98 to 1.11 log units. However, $f_{\textrm{mean}}$ appears to lead to the best performance (RMSE = 0.98 and
$R^2$ = 0.67). In contrast, the RMSE and $R^2$ values for Y-scrambled\cite{Rucker:2007aa} models
using the different aggregation functions ranged from 1.58 to 1.61 log units and 0.003 to 0.01, respectively.

@@

We observed that the performance of models based on geometric mean aggregation function did not
differ substantially from those built with the arithmetic mean. This was also true of the other
datasets and so the following discussion omits the results obtained when using
$f_{\textrm{geom}}$. In fact, a paired $t$-test between the predicted SALI values using models built
with the different aggregation functions (i.e., $f_{\textrm{mean}}$ vs $f_{\textrm{absdiff}}$, etc.)
indicated that the differences between the sets of predictions was statistically insignificant (all
$p$-values $> 0.1$). Fig.~\ref{fig:rfmodel-chembl} summarizes the models built on the ChEMBL
datasets using $f_{\textrm{diff}}$ and $f_{\textrm{mean}}$ as the aggregation functions. Table
\ref{tab:rfmodel-results} reports the model statistics.  While the differences in model performance
do not vary significantly with the aggregation function used, we see that for the Costanzo and Kalla
datasets, $f_{\textrm{diff}}$ leads to a slightly better model in terms of RMSE and $R^2$ than when
using $f_{\textrm{mean}}$.

Given this observation, we arbitrarily chose $f_{\textrm{mean}}$ as the aggregation function
for subsequent analyses. Figs.~\ref{fig:costanzo-pred}, \ref{fig:kalla-pred} and \ref{fig:dai-pred}
summarize the performance of the models built using the reported activity data and the pairwise SALI
values for the Costanzo, Kalla and Dai datasets, respectively. As with the Cavalli dataset, the SALI
models exhibit more variance for small SALI values (i.e., less significant cliffs) but appear to
perform better at the higher end of the SALI values. However in all three cases, the most
significant cliffs exhibit significant variance. In comparison, the models developed on the reported
activity values are relatively poor. This can be partially ascribed to the small sizes of the
dataset. It should be note that this methodology is not aimed at modeling the original, usually
small, SAR dataset. It is well known (and evident from the current analyses) that small datasets do
not lead to very robust or reliable machine learning models. Instead, the methodology focuses on the
\emph{pairwise} dataset -- which are usually large enough for a statistically significant model.

>>>>

Given the independent and dependent variables for the pairwise dataset, we can now proceed to model
development. For this study we focused on the use of random forest models\cite{Breiman:1984aa}. This
was motivated by the fact that such models tend not to overfit the training data (given a
sufficiently large number of trees) and the fact that the algorithm implicitly performs feature
selection. As a result, this allows us to forgo an explicit feature selection step and work directly
with the descriptor pool (after removal of correlated and constant descriptors). Furthermore, there
is no reason, \emph{a priori} to assume that the underlying SAR's are linear. A random forest model,
being an algorithmic approach\cite{Breiman:2001nx} (as opposed to a distributional one such as
linear regression) makes no such assumptions. We employed the implementation of random forest from
the \texttt{randomForest} package, in R 2.11.0\cite{r}. The models were built using 500 trees, and
randomly sampled the descriptor pool for a given dataset, using $\sqrt(N)$ descriptors at a time,
where $N$ is the number of descriptors in the pool. By default, the method builds individual trees
using 63\% of the dataset and tested on the remainder (the so called out-of-bag data). The final
predicted values for the entire dataset are obtained by averaging the predictions for each
observation from from all 500 trees. We initially employed a
training/test set split (80\% of the dataset was assigned to the training set) and these results are
reported in Table \ref{tab:data}.

@@

\section{Results}
\label{sec:applications}
We first consider the application of the SALI prediction methodology to the Cavalli dataset. We
developed three random forest models, using the three aggregation descriptor aggregation functions
described in Section \ref{sec:methodology}. Fig.~\ref{fig:rfmodel-cavalli} presents the predicted
versus observed SALI values from the three models and Table \ref{tab:rfmodel-results} summarizes the
performance metrics for the three models. Overall the different aggregation functions do not differ
dramatically in terms of the final model performance as root mean square error (RMSE) on the test
set ranges from 1.01 to 1.07 log units. However, $f_{\textrm{mean}}$ appears to lead to the best
performance (RMSE = 1.01 and $R^2$ = 0.58). In contrast, the RMSE and $R^2$ values for
Y-scrambled\cite{Rucker:2007aa} models using the different aggregation functions ranged from 1.58 to
1.61 log units and 0.003 to 0.01, respectively.

@@

We observed that the performance of models based on geometric mean aggregation function did not
differ substantially from those built with the arithmetic mean. This was also true of the other
datasets and so the following discussion omits the results obtained when using
$f_{\textrm{geom}}$. In fact, a paired $t$-test between the predicted (from the training set) SALI
values using models built with the different aggregation functions (i.e., $f_{\textrm{mean}}$ vs
$f_{\textrm{absdiff}}$, etc.)  indicated that the differences between the sets of predictions was
statistically insignificant (all $p$-values $> 0.1$). Fig.~\ref{fig:rfmodel-chembl} summarizes the
models built on the ChEMBL datasets using $f_{\textrm{diff}}$ and $f_{\textrm{mean}}$ as the
aggregation functions. Table \ref{tab:rfmodel-results} reports the model statistics.  While the
differences in model performance do not vary significantly with the aggregation function used, we
see that for all the ChEMBL datasets, $f_{\textrm{diff}}$ leads to a slightly better model
in terms of RMSE and $R^2$ than when using $f_{\textrm{mean}}$.

Given this observation, we arbitrarily chose $f_{\textrm{diff}}$ as the aggregation function
for subsequent analyses. Figs.~\ref{fig:costanzo-pred}, \ref{fig:kalla-pred} and \ref{fig:dai-pred}
summarize the performance of the models built using the reported activity data and the pairwise SALI
values for the Costanzo, Kalla and Dai datasets, respectively. As with the Cavalli dataset, the SALI
models exhibit more variance for small SALI values (i.e., less significant cliffs) but appear to
perform better at the higher end of the SALI values. However in all three cases, the most
significant cliffs exhibit significant variance. In comparison, the models developed on the reported
activity values are relatively poor. This can be partially ascribed to the small sizes of the
dataset. It should be note that this methodology is not aimed at modeling the original, usually
small, SAR dataset. It is well known (and evident from the current analyses) that small datasets do
not lead to very robust or reliable machine learning models. Instead, the methodology focuses on the
\emph{pairwise} dataset -- which are usually large enough for a statistically significant model.
